\begin{thebibliography}{\hspace{5mm}}
%
\bibitem[AzDi]{AzDi} Azpeitia, A.G., and Dickinson, D.J.:
{\sl A decision rule in the simplex method thats avoid cycling.}
Numer. Math.. 6 (1964), 329-331.
%
\bibitem[Bau]{Bau} Baumgarten, Tobias.: {\sl Analyse und Vergleich numerischer
Verfahren der linearen Optimierung.} Diplomarbeit. Stuttgart 1996.
%
\bibitem[BaGo]{BaGo} Bartels, R.H., and Golub G.H.: {\sl
The simplex method of lonear programmming using LU decomposition.}
CACM 12 (1969), 266-268.
%
\bibitem[Ba]{Ba} Bartels, R.H.: {\sl A stabilization of simplex method.}
Numer. Math. 16 (1971), 414-434.
%
\bibitem[Be]{Be} Beer, K.: {\sl L”sung groáer linearer Optimierungsaufgaben.}
Berlin: Deutscher Verlag Wiss. 1977.
%
\bibitem[BeRi]{BeRi} Best, M.J., and Ritter, K.: {\sl Linear Programming.}
Englewood Cliffs, N.Y.: Prentice Hall 1985.
%
\bibitem[BeRi1]{BeRi1} Best, M.J., and Ritter, K.: {\sl Private
communication.} 1985.
%
\bibitem[Bla]{Bla} Bland, R.G.: {\sl New finite pivot rules for simplex
method.} Math. Operation Research 2 (1977), 103-107.
%
\bibitem[ColWe]{ColWe} Collatz, L., and Wetterling, W.: {\sl
Optimierungsaufgaben.}\\
Berlin--Heidelberg--New York: Springer 1966.
%
\bibitem[Dan]{Dan} Dantzig, G.B.: {\sl	Linear Programming and Extensions.}
Princeton: Princeton University Press 1963.
%
\bibitem[FleMat]{FleMat} Fletcher, R., and Matthews S.P.: {\sl Stable
modification of explizit LU factors for simplex updates.}
Math. Programming 29/30 (1984), 261-284.
%
\bibitem[FoTo]{FoTo} Forrest, J.J., and Tomlin, J.A.: {\sl Updating triangular
factors of the basis from simplex method.}
Math. Programming 2 (1972), 263-278.
%
\bibitem[Gro]{Gro} Groámann, Ch., and Terno, J.: {\sl Numerik der Optimierung}
Stuttgart: Teubner 1993.
%
\bibitem[Las]{Las} Lasdon, L.S.: {\sl Optimization Theory for Large Systems.}
London: Macmillan 1970.
%
\bibitem[Naz]{Naz} Nazareth, J.L.: {\sl Computer Solutions of Linear
Programs.} New York--Oxford: Oxford University Press 1987.
%
\bibitem[Pad]{Pad} Padberg, M.: {\sl Linear Optimization and Extensions.}
Berlin--Heidelberg--New York: Springer--Verlag 1995.
%
\bibitem[Schr]{Schr} Schrijver, A.: {\sl Theory of Linear and Integer
Programming.}
New York: Wiley 1986.
%
\bibitem[Spe]{Spe} Spellucci, P.: {\sl Numerische Verfahren der Nichtlinearen
Optimierung.} Basel--Boston--Berlin: Birkh„user 1992.
%
\end{thebibliography}
