%-- PROSEMINAR_LISTE ------------------------------------------------
\documentclass[12pt,a4paper]{article}
\input aaformat
\begin{document}
%\thispagestyle{empty}
\bc Arbeiten "uber Gitter in\\
    Int. J. Numer. Meth. Eng. (nach Ersch.-jahr)
\ec
%
%
%\begin{tabbing}
%1. \hspace{3mm} \= {\sc Gateaux-} {\sl und} {\sc Frechet}{\sl-Ableitungen.}\\
%   \> \cite{Ort}, \S \S  $\ $ 6.1 - 6.3; \cite{Lue}, \S \S $ \ $ 7.2, 7.3;
%      \cite{Kir}, \S $\ $ 4. \\[2mm]
%
%\end{tabbing}
\begin{thebibliography}{\hspace{5mm}}
%
\bibitem[Bor3]{Bor3} Borouchaki, H., and Frey, P.J.: {\sl Adaptive
triangular-quadrilateral mesh generation.}
41, 915-934.
%
\bibitem[Bor97]{Bor97} Borouchaki, H., und George, P.L.:  {\sl Aspects of 2-D
Delaunay mesh generation.}
40 (1997), 1957-1975.
%
\bibitem[Tsay]{Tsay} Tsay, T.K.: {\sl Numerical grid generation of an irregular
region.}
40, 343-356.
%
\bibitem[Riv3]{Riv3} Rivara, M.C.: {\sl New longest-edge algorithms for the
refinement and/or improvement of unstructured triangulations.}
40, 3313-3324.
%
\bibitem[Riv2]{Riv2} Rivara,M.-C., und Inostroza, P.: {\sl
Using longest-side bisection techniques for the automatic refinement of
Delaunay triangulations.}
40 (1997), 581-597.
%
\bibitem[Bor96]{Bor96} Borouchaki, H, et al.:
{\sl Optimal Delaunay point insertion.}
39 (1996), 3407-3437.
%
\bibitem[Dra]{Dra} Drake, R.: {\sl A method for dynamic mesh adaption.}
39,939-950.
%
\bibitem[Zav]{Zav} Zavattieri, P.D.: {\sl Optimization strategies
in unstructured domains.}
39, 2055-2072.
%
\bibitem[Rem]{Rem} Remondini, L., et al.: {\sl A new concept in
two-dimensional auto-adaptive mesh generation.}
37, 2841-2856.
%
\bibitem[Sub]{Sub} Subramanian, G., et al.: {\sl Robust boundary triangulation
and Delaunay triangulation of arbitrary planar domains.}
37 (1994), 1779-1789.
%
\bibitem[Fang]{Fang} Fang, J. et al.: {\sl Sweepline Algorithm for
unstructured-grid  generation on two-dimensional non-convex domains.}
36, 2761-2778.
%
\bibitem[Koh]{Koh} Kohli, H.S., and Carey, G.F.: {\sl Shape optimization using
adaptive shape refinement.}
36, 2435-2452.
%
\bibitem[Nam]{Nam} Nambiar, R.V., et al.: {\sl An algorithm for adaptive
refinement of triangular finite element meshes.}
36, 499-510.
%
\bibitem[Lee]{Lee} Lee, C.K., and Lo, S.H.:  {\sl An automatic adaptive
refinement finite element procedure for 2Delastostatic analysis.}
35, 1967-1990.
%
\bibitem[John2]{John2} Johnston, B.P., and Sullivan, J.M.: {\sl
Fully automatic two dimensional mesh generation using normal offsetting.}
33, 425-443.
%
\bibitem[Blae]{Blae} Blaecker, T.D., and Stephenson, M.B.: {\sl
Paving. A new approach to automated quadrilateral mesh generation.}
32, 811-848.
%
\bibitem[Zhu]{Zhu} Zhu, J.Z., et al.: {\sl A new approach to the development of
automatic quadrilateral mesh generation.}
32, 849-866.
%
\bibitem[Zien]{Zien} Zienkiewicz, O.C. and Zhu, J.Z.: {\sl Adaptivity and Mehs
generation.}
32, 783-810.
%
\bibitem[Sez]{Sez} Sezer, L., and Zeid, I.: {\sl Automatic
quadrilateral/triangular free-form mesh generation for planar regions.}
32, 1441- 1484.
%
\bibitem[Frey]3{Frey3} Frey, M.H., and Field, D.A.: {\sl Mesh relaxation: A new
technique for improving triangulations.}
31, 1121-1134.
%
\bibitem[John1]{John1} Johnston, B.P., et al.: {\sl Automatic conversion of
triangular finite element meshes to quadrilateral elements.}
31, 67-84.
%
\bibitem[Lo3]{Lo3} Lo, S.H.: {\sl Automatic mesh generation and adaption by
using contours.}
31, 689-708.
%
\bibitem[Del]{Del} Deljouie-Rakhshandeh, K.: {\sl An approach to the generation
of triangular grids possessing few obtuse triangles.}
29, 1299-1322.
%
\bibitem[Jin]{Jin} Jin, H., and Wiberg, N.E.: {\sl Two-dimensional mesh
generation, adaptive remeshing and refinement.}
29, 1501-1526.
%
\bibitem[Tal]{Tal} Talbert, J.A. and Parkinson, a.R.: {\sl Development of an
automatic, two-dimensional finite element mesh generator using quadrilateral
elements and Bezier curve boundary definition.}
29, 1589-1594
%
\bibitem[Che]{Che} Cheng, F. et al.: {\sl A parallel mesh generation algorithm
based on the vertex label assignement scheme.}
28, 1429-1448.
%
\bibitem[Lo2]{Lo2} Lo, S.H.: {\sl Delauny triangulation of non-convex planar
domains.}
28, 2695-2708.
%
\bibitem[Bae]{Bae} Baemann, P.L., et al.  :  {\sl Robust, geomatrically based,
automatic two-dimensional mesh generation.}
24, 1043-1078.
%
\bibitem[Frey2]{Frey2} Frey, M.H.: {\sl Selective refinement: A new strategy
for automatic node placement in graded triangular meshes.}
24 (1987), 2183-2200.
%
\bibitem[Riv1]{Riv1} Rivara, M.-C.: {\sl A grid generation based on 4-triangles
conforming mesh-refinement algorithms.}
24 (1987), 1343-1354.
%
\bibitem[Joe]{Joe} Joe, B., and Simpson, R.B.; {\sl Triangular meshes for
regions of complicated shape.}
23, 751-778.
%
\bibitem[Lo1]{Lo1} Lo, S.H.: {\sl A
new mesh generation scheme for arbitrary planar domains.}
21 (1985), 1403-1426.
%
\bibitem[Mos]{Mos} Moscardini, A.O., et al.: {\sl AGTHOM--Automatic generation
of triangular and higher oder meshes.}
19, 1331-1354.
%
\bibitem[Klei]{Klei} Kleinstreuer, C. und Holdeman, J.T.: {\sl A triangular
finite element mesh generator for fluid dynamic systems of arbitrary
geometry.}
15, 1325-1334.
%
\bibitem[Ima]{Ima} Imafuku, I., et al.: {\sl A generalized automatic
mesh generation scheme for finite element methods.} 15, 713-732.
%
\bibitem[Cav]{Cav} Cavendish, J.C.: {\sl Triangulation of arbitrary planar
domains for the finite element method.} 8, 679-696.
%
\bibitem[Fre1]{Fre1} Frederick, C.O. et
al.: {\sl Two-dimensional automatic mesh generation for dtructural analysis.}
2, 133-144.
%
\end{thebibliography}
%
\end{document}
