\bibliographystyle{alpha}
\begin{thebibliography}{WWW99}
  
\bibitem{hibernatebook}
  King, G., Bauer, C.
  \newblock \emph{Hibernate In Action}.
  \newblock Manning Publications, Greenwinch, 2005.

\bibitem{designpatterns}
  Gamma, E., Helm, R., Johnson, R. \& Vlissides, J.
  \newblock \emph{Design Patterns: Elements of Reusable Object-Oriented Software}.
  \newblock Addison Wesley, Reading Mass., 1995.

\bibitem{ooconstruction}
  Meyer, B.
  \newblock \emph{Object Oriented Software Construction, Second Edition}.
  \newblock Prentice Hall, New Jersey, USA, 1997.

\bibitem{jmsbook}
  Chappel, D., Monson-Haeffel, R.
  \newblock \emph{Java Message Service}.
  \newblock O'Reilly, Sebastapol CA, 2001.

\bibitem{jbossadmin}
  JBoss, Inc.
  \newblock \emph{JBoss Admin Development Guide - JBoss 3.2.6}.
  \newblock JBoss, Inc., 2004.

\bibitem{jbossclustering}
  Burke, B.
  \newblock \emph{JBoss Clustering}.
  \newblock JBoss Group, Atlanta, GA, 2003.
  
\bibitem{perfbook}
  Looseley, C., Douglas, F.
  \newblock \emph{High-Performance Client/Server}.
  \newblock Wiley Computer Publishing, 1997.

\bibitem{tcpip}
  Hunt, C.
  \newblock \emph{TCP/IP Network Administration}.
  \newblock O'Reilly, 1998

\bibitem{codd}
  Codd, E. F.
  \newblock \emph{A Relational Model for Large Shared Data Banks}.
  \newblock CACM, 1:6, 1970
 
\bibitem{dbfund}
  Elmasri, R. Navathe, B. S.
  \newblock \emph{Fundemantals of Database Systems, Second Edition}.
  \newblock Addision-Wesley, 1994

\bibitem{manmonth}
  Brooks, F. P. Jr.
  \newblock \emph{The Mythical Man-Month, 1995 Anniversary Edition}.
  \newblock Addision-Wesley, 1995
  
\bibitem{complexity}
  Lewin, R.
  \newblock \emph{Complexity, Life at the Edge of Chaos, Second Edition}.
  \newblock Phoenix, 2001
   
\bibitem{nixon}
  Nixon, R.
  \newblock \emph{Six Crises}.
  \newblock Touchstone Press, New York, 1990
   
\bibitem{torvalds}
  Torvalds, L., Diamond, D.
  \newblock \emph{Yanlýzca Eðlenmek Ýçin (Orijinal: Just For Fun)}.
  \newblock Bilgi Yayýnevi, 2005, Orijinal: Harper Business, New York, 2001
   
\end{thebibliography}
  
